An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals

### 2022-02-09 analysis/paper/paper.Rmd
Almost done with edits for reviewers!

Here we present a method for generating a BCG from bio-logger accelerometry. We validated our method with a simultaneously recorded ECG on an adult killer whale in managed care (*Orcinus orca*) and applied it to detect heartrate in a blue whale. The relative orientation of the tag on the body is often uncertain when bio-loggers are deployed in the wild [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on cranio-caudal acceleration, we also generated a three-dimensional BCG, which we expected would be more robust in a field setting. Specifically, we tested three hypotheses to validate our method. First, a one-dimensional BCG would, in a controlled setting, produce instantaneous heartrates that are statistically equivalent to ECG instantaneous heartrates. Second, a three-dimensional BCG would, in a field setting, produce a more robust signal than a one-dimensional BCG. Third, BCG-derived heartrates would increase during the latter phases of dives, consistent with the progressive increase in heartrate routinely observed prior to and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].

@@

Here we presented a ballistocardiogram method for detecting resting apneic heartrate in cetaceans using accelerometers. We validated the method in a controlled setting with simultaneous ECG and in a field setting by confirming expected physiological patterns. As accelerometer tags have been deployed on many cetacean species for multiple decades, this method may be applied to mine existing datasets and better understand how heartrate scales with body size and other biological factors. It may also provide additional data for conservation physiology applications. For example, BCGs extracted from gliding phases before and after controlled sonar exposure experiments could quantify the physiological response to anthropogenic disturbance [@southall2019]. Even as the field of physio-logging progresses with new hardware innovations, this method demonstrates that computational advances can derive new insights from traditional sensors.

>>>>

Here we present a method for generating a BCG from bio-logger cranio-caudal acceleration. We validated our method with a simultaneously recorded ECG on an adult killer whale in managed care (*Orcinus orca*) and applied it to detect heartrate in a blue whale. The relative orientation of a tag on a cetacean's body is often uncertain when bio-loggers are deployed in the wild [@johnsonDigitalAcousticRecording2003], so isolating acceleration along the cranio-caudal axis is subject to error. Therefore, we also compared a tri-axial BCG to the cranio-caudal BCG. Specifically, we tested three hypotheses to validate our method. First, a cranio-caudal (1D) BCG would, in a controlled setting, produce instantaneous heartrates that are statistically equivalent to ECG instantaneous heartrates. Second, a tri-axial (3D) BCG would, in a field setting, produce a more robust signal than a 1D BCG. Third, BCG-derived heartrates would increase during the latter phases of dives, consistent with the progressive increase in heartrate routinely observed prior to and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].

@@

Here we presented a ballistocardiogram method for detecting resting apneic heartrate in cetaceans using accelerometers. We validated the method in a controlled setting with simultaneous ECG and in a field setting by confirming expected physiological patterns. As accelerometer tags have been deployed on many cetacean species for multiple decades, this method may be applied to mine existing datasets and better understand how heartrate scales with body size and other biological factors. It may also provide additional data for conservation physiology applications. For example, BCGs extracted from gliding phases before and after controlled sonar exposure experiments could quantify the physiological response to anthropogenic disturbance [@southall2019]. Even as the field of physio-logging progresses with new hardware innovations [@williams2021; @fahlman2021; @hawkesIntroductionThemeIssue2021], this method demonstrates that computational advances can derive new insights from traditional sensors.
